---
title: "Extending {ggformula}"
author: "Randall Pruim"
date: last-modified
format: html
engine: knitr
toc: true
vignette: >
  %\VignetteIndexEntry{Extending ggformula}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  engine: knitr
  eval: true
  echo: true
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r setup, include = FALSE}
have_packages <-
  require(ggformula) &&
  require(ggplot2) &&
  requireNamespace("ggrepel") &&
  requireNamespace("ggdist")

knitr::opts_chunk$set(
  fig.show = "asis",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "60%",
  eval = have_packages
)
theme_set(theme_light())
```

```{r missing-packages-warning, eval = !have_packages, echo = FALSE, results = 'asis'}
cat(
  "
## Warning: Missing packages

Because one or more of `ggformula`, `ggplot2`, `ggrepel`, and `ggdist`
appears to be missing, this vignette is compiling without executing any
code.

"
)
```

Every `gf_*` function in `ggformula` -- `gf_point()`, `gf_boxplot()`,
`gf_ribbon()`, and around a hundred more -- is built by a single function
factory, `layer_factory()`. This vignette explains how that factory works
well enough that you can use it to add your own `gf_*` wrapper around a
`geom`/`stat` from another package, and walks through two worked examples:
labeling points without overlap using [ggrepel](https://ggrepel.slowkow.com/),
and half-eye/raincloud-style distribution plots using
[ggdist](https://mjskay.github.io/ggdist/).

You do not need to be a `ggformula` developer to follow along -- this is
meant for anyone who wants a formula-based wrapper around a geom or stat
that `ggformula` doesn't already provide.

## How `layer_factory()` works

Every `gf_*` function is created by calling `layer_factory()` with a small
set of arguments describing the geom/stat it wraps. Here, roughly, is how
`gf_point()` is defined inside `ggformula` itself:

```r
gf_point <-
  layer_factory(
    geom = "point",
    aes_form = y ~ x,   # this is the default value
    extras = alist(
      alpha = , color = , size = , shape = , fill = , group = , stroke =
    )
  )
```

The important arguments are:

- **`geom`** / **`stat`** / **`position`** -- the (usually low-level)
  `ggplot2` geom, stat, and position to use, given either as strings (e.g.
  `"point"`, which resolves to `GeomPoint`) or as actual objects/functions.
- **`aes_form`** -- a formula, or a list of formulas, describing the
  formula shape(s) the function accepts. It defaults to `y ~ x`, which is
  why `gf_point(mpg ~ wt, data = mtcars)` maps `wt` to `x` and `mpg` to
  `y`. Some functions need more than one shape -- `gf_ribbon()`, for
  example, uses `list(ymin + ymax ~ x, y ~ xmin + xmax)`.
- **`extras`** -- an `alist()` of additional arguments the function should
  accept, with defaults where relevant (an empty default, as in
  `alpha = `, means "no default; only include this if the user supplies
  it"). Each of these can be *set* (`color = "red"`) or *mapped*
  (`color = ~species`) when the function is called.
- **`layer_fun`** -- the function ultimately used to build the layer.
  This defaults to `ggplot2::layer()`, which is the right choice when
  `geom`/`stat` name registered `Geom*`/`Stat*` ggproto objects. If you
  instead want to reuse an existing high-level constructor function (like
  `ggrepel::geom_text_repel()` or `ggdist::stat_halfeye()`) rather than
  re-implementing its logic, you point `layer_fun` at that function
  instead -- more on this below.

You can inspect the arguments any existing `gf_*` function was built with
via `ggformula_spec()`:

```{r inspect-spec}
str(ggformula_spec(gf_ribbon))
```

## Two patterns for wrapping a new geom or stat

There are two ways to plug an extension package's geom or stat into
`layer_factory()`, and which one you want depends on how that package
exposes its functionality.

### Pattern 1: point at the registered ggproto object by name

If the extension package registers a `Stat*`/`Geom*` ggproto object (most
`ggplot2` extension packages do), you can use the low-level
`geom =`/`stat =` arguments.  This is how `ggformula` defines `gf_sina()`
as a wrapper around `ggforce`'s sina-plot jitter, for example:

```r
gf_sina <-
  layer_factory(
    pre = {
      if (!requireNamespace("ggforce", quietly = TRUE)) {
        stop("The ggforce package is required.  Please install and try again.")
      }
      if (!"package:ggforce" %in% search()) {
        stop(
          "To use gf_sina(), the ggforce package must be loaded.\n    Try, for example, `library(ggforce)`."
        )
      }
    },
    geom = "point",
    stat = "sina",
    position = "identity",
    extras = alist(alpha = , color = , size = , fill = , group = )
  )
```

The `pre` code isn't strictly required, but it's good practice: because
this pattern resolves `stat = "sina"` to `StatSina` by searching the
*attached* packages (not just installed ones), the extension package must
be loaded with `library()`, not merely installed, for this to work. The
`pre` block gives a clear error message instead of a confusing one if a
user forgets.

### Pattern 2: wrap an existing constructor function

Many extension packages expose their functionality only (or best) through
a full constructor function -- 
like `ggplot2::geom_abline()`, `ggrepel::geom_text_repel()`, 
or `ggdist::stat_halfeye()` -- 
rather than a bare ggproto object you're expected to assemble yourself. 
In that case, point `layer_fun` at that function directly, and set 
`geom`/`stat` to a string with the same name (no `geom_`/`stat_` prefix) 
purely so that `layer_factory()` can look up that function's own formals 
to figure out which extra arguments to allow:

```r
gf_abline <-
  layer_factory(
    geom = "abline",
    aes_form = NULL,
    extras = 
      alist(slope = , intercept = , color = , linetype = , linewidth = , alpha = ),
    inherit.aes = FALSE,
    data = NA,
    layer_fun = rlang::quo(ggplot2::geom_abline)
  )
```

This is the pattern used for both examples below.

## Example: labeling points without overlap with {ggrepel}

[ggrepel](https://ggrepel.slowkow.com/) provides `geom_text_repel()` and
`geom_label_repel()`, drop-in replacements for `ggplot2::geom_text()`/
`geom_label()` that nudge overlapping labels apart. `ggformula` already
has `gf_text()` and `gf_label()`; here's a `gf_text_repel()` built the
same way, but pointed at `ggrepel::geom_text_repel()`.

```{r gf-text-repel-definition}
library(ggrepel)

gf_text_repel <-
  layer_factory(
    geom = "text_repel",
    layer_fun = rlang::quo(ggrepel::geom_text_repel),
    extras = alist(
      label = ,
      alpha = ,
      color = ,
      size = ,
      fontface = ,
      family = ,
      box.padding = 0.25,
      point.padding = 1e-06,
      min.segment.length = 0.5,
      max.overlaps = 10,
      nudge_x = 0,
      nudge_y = 0,
      seed = NA,
      direction = "both"
    )
  )
```

A few things to note:

- `geom = "text_repel"` doesn't correspond to a real `Geom*` object; it's
  only used to fetch `formals(geom_text_repel)` so those argument names
  (`box.padding`, `max.overlaps`, etc.) are recognized automatically, in
  addition to the ones listed explicitly in `extras`.
- Because of this, `layer_fun` is required here and tells `ggformula` where to
locate the function used to create a plot layer.
- `aes_form` was left at its default, `y ~ x`, which is exactly what we
  want here.
- `label` is listed in `extras` with no default, the same way `gf_text()`
  handles it, so it can be set (`label = "winner"`) or mapped
  (`label = ~name`).

Using it looks just like using `gf_text()`:

```{r gf-text-repel-example}
df <- data.frame(
  x = c(1, 2, 3, 2.9, 5),
  y = c(3, 1, 4, 4.1, 5),
  name = c("apple", "banana", "cherry", "date", "elderberry")
)

gf_point(y ~ x, data = df) |>
  gf_text_repel(y ~ x, label = ~name, seed = 1234)
```

Compare that to `gf_text()`, which lets the labels overlap:

```{r gf-text-example}
gf_point(y ~ x, data = df) |>
  gf_text(y ~ x, label = ~name, nudge_y = 0.3)
```

## Example: half-eye plots with {ggdist}

[ggdist](https://mjskay.github.io/ggdist/) provides "raincloud"-style
`stat_halfeye()`, which draws a half-violin density alongside a point
estimate and one or more uncertainty intervals -- a richer alternative to
`gf_violin()`/`gf_boxplot()`. `stat_halfeye()` is itself a high-level
constructor (its default `geom` is `"slabinterval"`), so this again uses
Pattern 2.

```{r gf-halfeye-definition}
library(ggdist)

gf_halfeye <-
  layer_factory(
    geom = "slabinterval",
    stat = "halfeye",
    layer_fun = rlang::quo(ggdist::stat_halfeye),
    extras = alist(
      fill = ,
      color = ,
      alpha = ,
      adjust = 1,
      point_interval = "median_qi",
      .width = c(0.66, 0.95),
      side = "top",
      justification = NULL
    )
  )
```

Setting `stat = "halfeye"` here means `layer_factory()` looks up
`formals(stat_halfeye)` itself (since `stat_halfeye()` is both the
constructor we're calling *and* the thing we're using to discover valid
arguments), which automatically permits arguments like `point_interval`,
`.width`, `density`, and `breaks` without having to list every one of them
in `extras`.

```{r gf-halfeye-example}
set.seed(202)
scores <- data.frame(
  section = rep(c("A", "B", "C"), each = 30),
  score = c(rnorm(30, 78, 6), rnorm(30, 82, 9), rnorm(30, 75, 5))
)

gf_halfeye(score ~ section, data = scores, fill = ~section, show.legend = FALSE)
```

`stat_halfeye()` sets its own default of `show.legend = c(size = FALSE)`
when called directly, to avoid an unwanted legend for its point-size
aesthetic; because `layer_factory()`-built functions always pass an
explicit `show.legend` through to `layer_fun` (`NA` by default), that
sensible default gets overridden. Passing `show.legend = FALSE` explicitly,
as above, avoids the stray legend entry.

## Wrapping interactive geoms from `ggiraph`

If you'd like an `_interactive` counterpart of your new function (for use
with `gf_girafe()`), you generally don't need to do anything extra:
`interactive_layer_factory()` builds one automatically from any function's
`ggformula_spec()`, as long as [ggiraph](https://davidgohel.github.io/ggiraph/)
provides an interactive version of the same geom (e.g.
`ggiraph::geom_text_repel_interactive()`). See `vignette("interactive-graphics")`
for more on interactive plots in general.

## Tips and things to watch for

- **Attach vs. install.** If you use Pattern 1 (a bare `geom =`/`stat =`
  name), the extension package must be *attached* with `library()`, not
  merely installed, because `ggplot2` resolves those names by searching
  attached namespaces. Pattern 2 (`layer_fun = `) doesn't have this
  restriction, since you're calling the extension's function directly.
- **`check.aes`.** By default, `layer_factory()` warns if you supply an
  aesthetic that isn't among the geom's/stat's known aesthetics. Set
  `check.aes = FALSE` if you need to pass an aesthetic that
  `layer_factory()` can't discover automatically (this is rarely
  necessary if you list the relevant names in `extras`).
- **`pre` for guardrails.** The `pre` argument lets you run arbitrary code
  before the layer is built -- useful for clear error messages when an
  optional dependency isn't attached (as in the `gf_sina()` example
  above), or for small argument-massaging steps like the `nudge_x`/
  `nudge_y` handling used internally by `gf_text()`.
- **Testing.** `ggformula`'s own test suite uses
  `vdiffr::expect_doppelganger()` (wrapped as `wrapped_expect_doppelganger()`
  internally) to catch unintended rendering changes; the same approach
  works well for testing a new wrapper you've written.

## Further reading

- `?layer_factory` for the full list of arguments.
- `?ggformula_spec` for introspecting existing `gf_*` functions.
- `vignette("ggformula")` for the base formula syntax that every `gf_*`
  function (including ones you build yourself) inherits for free.
