---
title: "Formula Interface for ggplot2"
author: "Daniel Kaplan and Randall Pruim"
date: "July, 2020"
output:  rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Formula Interface for ggplot2}
---

```{r setup, include = FALSE}
have_packages <-
  require(ggformula) &&
  require(dplyr) &&
  require(ggplot2) &&
  require(mosaicData) &&
  require(maps) &&
  require(modeldata) &&
  requireNamespace("mosaic") 

have_extra_packages <- 
  require(maps) && require(sf) && require(purrr)  

knitr::opts_chunk$set(
  fig.show = "asis",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  out.width = "60%",
  eval = have_packages
)
theme_set(theme_light())
```

```{r, eval = ! have_packages, echo = FALSE, results = 'asis'}
cat(
"
## Warning: Missing packages

Because one or more of 
`ggformula`, `ggplot2`, `dplyr`, `mosaic`, `mosaicData`, `modeldata`,
and `maps`, appears to be missing, this vignette is compiling
without executing any code. 

"
)
```

# Formula-driven graphics

There are several excellent graphics packages provided for R. The
`ggformula` package currently builds on one of them, `ggplot2`, but
provides a very different user interface for creating plots. The interface
is based on formulas (much like the `lattice` interface) and the use of 
the chaining operator (`%>%`) to build more complex graphics from simpler 
components.


The `ggformula` graphics were designed with several user groups in mind:

  * beginners who want to get started quickly and may find the syntax of `ggplot2()` 
  a bit offputting,
  
  * those familiar with `lattice` graphics, but wanting to be
  able to easily create multilayered plots,
  
  * those who prefer a formula interface, perhaps because it is 
  familiar from use with functions like `lm()` or from use of 
  the `mosaic` package for numerical summaries.
  

## The basic formula template

The basic template for creating a plot with `ggformula` is

```{r, plottype, eval = FALSE}
gf_plottype(formula, data = mydata)
```

or, equivalently, 

```{r, plottype2, eval = FALSE}
mydata %>% gf_plottype(formula)
```

where 

  * `plottype` describes the type of plot (layer) desired (points, lines, a histogram,
etc., etc.), 

  * `mydata` is a data frame containing the variables used in the plot, and 
  
  * `formula` describes how/where those variables are used.

For example, in a bivariate plot, `formula` will take the form `y ~ x`, where `y` is the 
name of a variable to be plotted on the y-axis and `x` is the name of a variable to 
be plotted on the x-axis.  (It is also possible to use expressions that can be evaluated
using variables in the data frame as well.)

The first form of the tempate is useful for simple plots or for multi-layered plots
where different layers use different data.  The second form is useful for multi-layered
plots or plots with many arguments.

Here is a simple example:
```{r, simple-example}
library(ggformula)
gf_point(mpg ~ hp, data = mtcars)
mtcars %>% gf_point(mpg ~ hp)
```


## Selecting the glyph type

The "kind of graphic" is specified by the name of the graphics function. All of
the `ggformula` data graphics functions have names starting with
`gf_`, which is intended to remind the user that they are formula-based
interfaces to `ggplot2`: `g` for `ggplot2` and `f` for "formula." 
Commonly used functions include

- `gf_point()` for scatter plots
- `gf_line()` for line plots (connecting dots in a scatter plot)
- `gf_density()` or `gf_dens()` or `gf_histogram()` or `gf_dhistogram()` or `gf_freqpoly()` to display distributions of a quantitative variable
- `gf_boxplot()` or `gf_violin()` for comparing distributions side-by-side
- `gf_counts()` for bar-graph style depictions of counts.
- `gf_bar()` for more general bar-graph style graphics

The function names generally match a corresponding function name from `ggplot2`,
although 

  * `gf_counts()` is a simplified special case of `geom_bar()`, 
  * `gf_dens()` is an alternative to `gf_density()` that displays the density plot 
  slightly differently 
  * `gf_dhistogram()` produces a density histogram rather than a count histogram.

Each of the `gf_` functions can create the coordinate axes and fill it in one
operation. (In `ggplot2` nomenclature, `gf_` functions create a frame and add a
layer, all in one operation.)  This is what happens for the first
`gf_` function in a chain.  For subsequent `gf_` functions, new layers are added,
each one "on top of" the previous layers.

## Attributes

Each of the marks in the plot is a *glyph*. Every glyph has graphical *attributes*
(called aesthetics in `ggplot2`) that tell where and how to draw the glyph. 
In the above plot, the obvious attributes are x- and y-position:  
We've told R to put `mpg` along the y-axis and `hp` along the x-asis, as is clear
from the plot.

But each point also has other attributes, including color, shape, size, stroke, fill, 
and alpha (transparency).  We didn't specify
those in our example, so `gf_point()` uses some default values for those --
in this case smallish black filled-in circles.

### Specifying attributes

In the `gf_` functions, you specify the non-position graphical attributes using 
additional arguments to the function.
Attributes can be **set** to a constant value
(e.g, set the color to "blue"; set the size to 2) 
or they can be **mapped** to a variable in the data or some 
expression involving the variables 
(e.g., map the color to `sex`, so sex determines the color groupings)

Attributes are set or mapped using additional arguments.

 * adding an argument of the form `attribute = value` **sets** `attribute` to `value`.
 * adding an argument of the form `attribute = ~ expression` **maps** `attribute` to `expression`
 
where `attribute` is one of `color`, `shape`, etc., `value` is a constant 
(e.g. `"red"` or `0.5`, as appropriate), and `expression`
may be some more general expression that can be computed using the variables in `data` 
(although often is is better to create a new variable in the data and to
use that variable instead of an on-the-fly calculation within the plot).


The following plot, for instance, 

 * We use `cyl` to determine the color and `carb` to determine the size of each
 dot.  Color and size are **mapped** to `cyl` and `carb`. 
 A legend is provided to show us how the mapping is being done.
 (Later, we can use scales to control precisely how the mapping is done -- 
 which colors and sizes are used to represent which values of `cyl` and `carb`.)  
 
 * We also **set** the transparency to 50%.  The gives the same value of `alpha` to
 all glyphs in this layer.
 
```{r, mapping-setting}
gf_point(mpg ~ hp, color = ~ cyl, size = ~ carb, alpha = 0.50, data = mtcars) 
```

### On-the-fly calculations

`ggformula` allows for on-the-fly calculations of attributes, although the default labeling 
of the plot is often better if we create a new variable in our data frame.  In the 
examples below, since there are only three values for `carb`, it is easier to read the 
graph if we tell R to treat `cyl` as a categorical variable by converting to a factor (or to 
a string).  Except for the labeling of the legend, these two plots are the same.
In the second example, we see how the ggformula works well with data tranformations
using `%>%`.

```{r, on-the-fly}
library(dplyr)
gf_point(mpg ~ hp,  color = ~ factor(cyl), size = ~ carb, alpha = 0.75, data = mtcars)
mtcars %>% 
  mutate(cylinders = factor(cyl)) %>% 
  gf_point(mpg ~ hp,  color = ~ cylinders, size = ~ carb, alpha = 0.75)
```


## "One-variable" plots

For some plots, we only have to specify the x-position because the y-position is calculated
from the x-values.  Histograms, densityplots, and frequency polygons are examples.
To illustrate, we'll use density plots, but the same ideas apply to 
`gf_histogram()`, and `gf_freqpolygon()` as well. 
*Note that in the one-variable density graphics, the variable whose density is to be calculated goes to the right of the tilde, in the position reserved for the x-axis variable.*

```{r penguins, warning=FALSE}
data(penguins, package = "modeldata")   
gf_density( ~ bill_length_mm, data = penguins)
gf_density( ~ bill_length_mm,  fill = ~ species,  alpha = 0.5, data = penguins)
# gf_dens() is similar, but there is no line at bottom/sides and the plot is not fillable
gf_dens( ~ bill_length_mm, color = ~ species,  alpha = 0.7, data = penguins)
# gf_dens2() is like gf_dens() but is fillable
gf_dens2( ~ bill_length_mm, fill = ~ species, data = penguins,
          color = "gray50", alpha = 0.4) 
```

Several of the plotting functions include additional arguments that do not modify
attributes of individual glyphs but control some other aspect of the plot.  In this
case, `adjust` can be used to increase or decrease the amount of smoothing.

```{r, dens}
# less smoothing
penguins %>% gf_dens( ~ bill_length_mm, color = ~ species, alpha = 0.7, adjust = 0.25)  
# more smoothing
penguins %>% gf_dens( ~ bill_length_mm, color = ~ species, alpha = 0.7, adjust = 4)     
```

## Position

When the `fill` or `color` or `group` aesthetics are mapped to a variable, the
default behavior is to lay the group-wise densities on top of one another. Other
behavior is also available by using the `position` argument. Using the value
`"stack"` causes the densities to be laid one on top of another.  This makes the
overall height of the stack represent the density across all groups, but it can be
difficult to discern the shape of the distribution for the individual groups. 
The value `"fill"` produces a conditional probability graphic.

```{r position, warning=FALSE}
penguins %>% 
  gf_density( ~ bill_length_mm, fill = ~ species, color = NA, position = "stack") 
penguins %>% 
  gf_density( ~ bill_length_mm, fill = ~ species, color = NA, position = "fill", 
              adjust = 2)
```

Similar commands can be constructed with `gf_histogram()` and `gf_freqpoly()`,
but note that `color`, not `fill`, is the active attribute for frequency
polygons.  It's also rarely good to overlay
histograms on top of one another -- better to use a density plot or a frequency 
polygon for that application.

## Faceting

The `ggplot2` system allows you to make subplots --- called "facets" --- based on
the values of one or two categorical variables. This is done by chaining
with  `gf_facet_grid()` or `gf_facet_wrap()`. These functions use formulas 
to specify which variable(s) are to be used for faceting.

```{r facets, warning=FALSE}
penguins %>% 
  gf_density_2d(bill_length_mm ~ bill_depth_mm) %>% 
  gf_facet_grid( ~ species)
# the dot here is a bit strange, but required to make a valid formula
penguins %>% 
  gf_density_2d(bill_length_mm ~ bill_depth_mm) %>% 
  gf_facet_grid( species ~ .)
penguins %>% 
  gf_density_2d(bill_length_mm ~ bill_depth_mm) %>% 
  gf_facet_grid(island ~ species)
penguins %>% 
  gf_density_2d_filled(bill_length_mm ~ bill_depth_mm, 
                       show.legend = FALSE) %>% 
  gf_facet_wrap(island ~ species)
```

An alternative syntax uses `|` to separate the faceting information from the main
part of the formula.  
Here is another example using our weather data.
The redundant use of the `y` and `color` attributes for 
temperature makes it easier to compare across facets.

```{r, facets02, out.width = "95%", fig.width = 6, fig.height = 3.5}
Weather %>% 
  gf_ribbon(low_temp + high_temp ~ date | city ~ year, alpha = 0.3) 

Weather %>% 
  gf_linerange(low_temp + high_temp ~ date | city ~ year, color = ~ avg_temp) %>%
  gf_refine(scale_colour_gradientn(colors = rev(rainbow(5))))
```

In this case, we should either not facet by year, or allows the x-scale to be freely adjusted in each column so that we don't have so much unnecessary white space.  We can do the latter
using the `scales` argument to `gf_facet_grid()`.

```{r, facets03, out.width = "95%", fig.width = 6, fig.height = 3.5}
Weather %>% 
  gf_ribbon(low_temp + high_temp ~ date | city ~ ., alpha = 0.3) 

Weather %>% 
  gf_linerange(low_temp + high_temp ~ date, color = ~ avg_temp) %>%
  gf_refine(scale_colour_gradientn(colors = rev(rainbow(5)))) %>%
  gf_facet_grid(city ~ year, scales = "free_x")
```

## Labels in plots

Usually plot labels reflect the expression mapped to a given attribute. But in the previous
example, our colorbar legend is labeled "avg temp (F)" even though the variable name
is `avg_temp`. This is because the `Weather` data set is *labeled*.

```{r}
get_labels(Weather)
```

ggformula will detect these labels and use them on our plots.  We can use `set_labels()`
to add custom labels to some or all of the variables in a data frame.^[labels]

[labels]: ggformula also detects labels created by functions in the expss and Hmisc packages.

```{r}
KF <- KidsFeet %>% 
  set_labels(
    length = "length (cm)", 
    width = "width (cm)",
    domhand = "dominant hand",
    biggerfoot = "bigger foot"
  )

KidsFeet %>% 
  gf_point(width ~ length, color = ~domhand)
KF %>%
  gf_point(width ~ length, color = ~domhand)

Penguins <- penguins %>%
  set_labels(
    species = "Species",
    island = "Island",
    bill_length_mm = "Bill Length (mm)",
    bill_depth_mm = "Bill Depth (mm)",
    flipper_length_mm = "Flipper Length (mm)",
    body_mass_g = "Body Mass (g)",
    sex = "Sex"
  )

Penguins %>%
  gf_point(bill_length_mm ~ bill_depth_mm, color = ~ species) 
```

By adding labels to the data, we can ensure consitent labeling across several plots.
But it is also possible to modify labels just for one plot.
This can be done in three ways:

* We can use arguments like `xlab`, `ylab`, and `title`;
* We can use the `gf_labs()` function (which has arguments like `x`, `y`, `title`, etc.);
* We can use the `gf_relabel()` function.

```{r}
KidsFeet %>%
  gf_point(width ~ length, color = ~domhand, xlab = "length (cm)", 
           title = "Length and width of kids' feet")

p <- KidsFeet %>% gf_point(width ~ length, color = ~domhand)

p %>% 
  gf_labs(x = "length (cm)", color = "dominant hand", title = "Some kids' feet")

p %>% 
  gf_relabel(length = "length (cm)", width = "width (cm)", domhand = "dominant hand")

# alternatively, we can use a list. 
# this is handy if we have several data sets with the same variables.
KFlabels <- 
  list(length = "length (cm)", width = "width (cm)", domhand = "dominant hand")
p %>% 
  gf_relabel(KFlabels)
```

Notice that `gf_labs()` and `gf_relabel()` take a different approaches to
labeling. `gf_labs()` refers to labels by their *role in the plot* whereas
`gf_relabel()` refers to variables in the data set and uses the new labels
wherever the variable appears.  Both are useful.  `gf_labs()` provides access to
additional labels (title, subtitle, caption, etc.), but it easier to mistakenly
create a plot with a label that doesn't match the true content of the plot.
`gf_relabel()` only deals with labels associated with variables, but the label
is guaranteed to match that variable no matter how the variable is used within
the plot.

## More 2-variable plots

### Using jitter and transparency to handle overlapping cases

Sometimes you have so many points in a scatter plot that they obscure one another. 
The `ggplot2` system provides two easy ways to deal with this: translucency and jittering.
Use `alpha` to make the points semi-translucent. 

Using `gf_jitter()` in place of `gf_point()` will move the plotted points to
reduce overlap.  Jitter and transparency can be used together as well.

```{r, jitter}
Penguins %>% 
  gf_point(bill_length_mm ~ species, alpha = 0.15)  
Penguins %>%
  gf_jitter(bill_length_mm ~ species, alpha = 0.35)
# width and heigh control the amount of horizontal and vertical jitter
Penguins %>%
  gf_jitter(bill_length_mm ~ species, alpha = 0.35, width = 0.2, height = 0)
```

### Box and Whisker plots

Box and whisker plots show the distribution of a quantitative variable as a
function of a categorical variable. The formula used in `gf_boxplot()` should
have the quantitative variable to the left of the tilde.  (To make horizontal 
boxplots using `ggplot2` you have to make vertical boxplots and then flip the 
coordinates with `coord_flip()`.)

```{r boxplot, warning = FALSE}
gf_boxplot(length ~ sex, color = "red", data = KF)
gf_boxplot(length ~ sex, color = ~biggerfoot , data = KF)
```


This plot may surprise you.

```{r, boxplot02}
Weather %>% 
  gf_boxplot(high_temp ~ month | city)
```

A single plot is placed at the at the mean value of `month` (for each city).
The warning message suggests that we need to tell R how to form the groups
when using a quantitative variable for `x`.
It suggests using the `group` aesthetic, and sometimes, this is 
just what we want.

```{r, boxplot03}
Weather %>%
  gf_boxplot(high_temp ~ month |  city, group = ~ month)
```

Alternatively, we could create a new categorical variable that puts the quantitative 
values into one of a few different bins.

### Alternatives to box and whisker plots.

`gf_violin()` and `gf_sina()` provide two alternatives to box and whisker plots.

```{r, violin-sina}
Penguins %>% gf_violin(bill_length_mm ~  species)
Penguins %>% gf_sina(bill_length_mm ~  species, alpha = 0.5)
Penguins %>%
  gf_violin(bill_length_mm ~  species) %>%
  gf_sina(bill_length_mm ~  species, alpha = 0.5, color = ~ island)
```


### 2-dimensional density plots

Two-dimensional plots of density also have both a left and right component to the formula.

```{r density2d-hex}
Penguins %>% gf_density_2d(bill_length_mm ~ bill_depth_mm | species)
Penguins %>% gf_density_2d_filled(bill_length_mm ~ bill_depth_mm | species)
Penguins %>% gf_hex(bill_length_mm ~ bill_depth_mm | species, binwidth = 1)
```

### Paths and lines

The `ggplot2` system offers two ways to connect points. `gf_line()` ignores the order of the points in the data, and draws the line going from left to right. `gf_path()` goes from point to point according to the order in the data. Both forms can use a `color` or `group` aesthetic to indicate which 
groups of points are connected.


<!-- The above are examples of *bad plots*. The viewer is unnecessarily distracted by -->
<!-- the zigs and zags in the connecting lines. It would be better to use -->
<!-- `gf_point()` here, but then you wouldn't see how `gf_line()` and `gf_path()` -->
<!-- work! -->

Here's an example where `gf_line()` is useful. 
We begin with a scatter plot showing the number of 
live births in the US for each day of 1978.

```{r, births01}
Births78 %>% gf_point(births ~ date)
```

Can this interesting pattern be explained by a weekday/weekend effect?  
Converting to a line plot and coloring by day of week 
highlights the pattern and makes it easy to spot the unusual days.

```{r, births03}
Births78 %>% gf_line(births ~ date, color = ~ wday)
```

The example above would look the same if we used `gf_path()` because the
data set is already sorted by date.  But in general, `gf_path()` and 
`gf_line()` produce different results.
In the plots below, the first connects days chronologically (because
the data are sorted by date) and the 
second in order of their low temperature.


```{r paths} 
Weather %>%
  filter(month == 5, year == 2017, city == "Chicago") %>%
  gf_path(high_temp ~ low_temp | city ~ year, color = ~ day) %>%
  gf_refine(scale_color_viridis_c(option = "C", begin = 0.2, end = 0.8))
Weather %>%
  filter(month == 5, year == 2017, city == "Chicago") %>%
  gf_line(high_temp ~ low_temp | city ~ year, color = ~ day) %>%
  gf_refine(scale_color_viridis_c(option = "C", begin = 0.2, end = 0.8))
```

## Plots with more than 2 positional attributes

Some layers require more than two attributes.  Typically this happens when the glyphs
of a layer are complex objects that could have been made using multiple layers, but 
belong together conceptually.  Examples include

  * `gf_pointrange()` -- plots a dot flanked by a line segment.  
  * `gf_linrange()` -- like `gf_pointrange()` but without the point.
  * `gf_errorbar()` -- vertical error bars.
  * `gf_errorbarh()` -- horizontal error bars.
  * `gf_ribbon()` -- a band between a line above and line below.
  
Often these are used to depict some sort of estimate of uncertainty in a
measurement or a prediction, but they can be used to represent any data of the
correct form.  Here we will use `gf_linerange()` and `gf_ribbon()` 
to indicate the high and low temperatures in New York for the first few months of 2013.

```{r, weather}
Temps <-
  Weather %>%
  filter(month <= 4, year <= 2016, city == "Chicago")

Temps %>% 
  gf_pointrange(avg_temp + low_temp + high_temp  ~ date, color = ~ avg_temp) %>% 
  gf_refine(scale_color_gradientn(colors = c("blue", "green", "orange", "red")))

Temps %>% 
  gf_ribbon(low_temp + high_temp  ~ date, color = "navy", alpha = 0.3)
```

## Positions and Stats

### Positions

`position_dodge()`, 
`position_jitter()`, and
`position_jitterdodge()` can be used to adjust the positions at which 
glyphs are placed.  Jittering adds some random noise and can be useful when
many observations have the same value.  Dodging moves groups of glyphs a fixed
difference to make it easier to distinguish the groups.

```{r, KidsFeet}
KidsFeet %>% 
  gf_point(length ~ sex, color = ~ domhand, 
           position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.4))
```

### Stats

A **stat** is a transformation that is applied to the data before 
a plot is generated.  Several of the plots we have seen have made use 
of stats.

 * `gf_histogram()` uses `stat_bin()` to bin the data and count the 
 number of observations in each bin.  It is equivalent to
 
```{r, stats}
HELPrct %>% gf_bar( ~ age, stat = "bin")
```


  * `gf_boxplot()` uses `stat_boxplot()` to compute the five-number summary on which
  the boxplot is based.
  
  * `gf_violin()`, `gf_density()`, and `gf_density2d()` use stats to compute a 
  sequence of points along an estimated density.  It is these points, and the not
  the raw data that are used to create the plot. `gf_sina()` uses the estimated 
  density to jitter points.
  
```{r boxplot-violin-sina}
gf_boxplot(age ~ substance, data = HELPrct)
gf_violin(age ~ substance, data = HELPrct) %>%
  gf_sina(alpha = 0.3)
```
  
There are also a number of stats that can be used to show a 
functional relationship between two variables.

* `gf_smooth()` plots a model fit using `lm()`, `glm()`, `gam()`, 
 `loess()`, `MASS::rlm()`.
 
* `gf_lm()` is `gf_smooth()` with the method set to `lm()` so it 
 plots a least squares regression line.
 
* `gf_spline()` plots a spline fit to the data.
 
```{r}
KidsFeet %>% 
  gf_point(length ~ width, color = ~ sex) %>%
  gf_lm()

mosaicData::Births %>%  
  gf_point(births ~ date, color = ~wday, alpha = 0.15) %>%
  gf_smooth(method = "loess")

mosaicData::Births %>%  
  gf_point(births ~ date, color = ~wday, alpha = 0.15) %>%
  gf_spline(spar = 1)
```

Confidence or prediction bands can be added to these as well.

```{r}
KidsFeet %>% 
  gf_point(length ~ width | sex, color = ~ sex) %>%
  gf_lm(interval = "prediction", fill = "red") %>%
  gf_lm(interval = "confidence", fill = "navy")
```

Mostly, the stats selected by default are just the ones you need. But sometimes
it is useful to select a different stat.  The `stat_summary()` and 
`stat_summary_bin()` stats are particularly useful in this respect.
These stats use a function to aggregate over unique values of `x` or over 
bins of `x` values and save the user needing to do that data transformation
manually. 

The default function applied in each group is `mean_se()`, which computes 
the mean (and the mean plus and minus one standard error)
This makes it simple to create an "interaction plot".

```{r, stats02}
KidsFeet %>% 
  gf_jitter(length ~ sex, color = ~ domhand, width = 0.1, height = 0) %>%
  gf_line(length ~ sex, color = ~ domhand, group = ~ domhand, stat = "summary")
Penguins %>% 
  gf_jitter(body_mass_g ~ sex, color = ~ species, width = 0.1, height = 0) %>%
  gf_line(body_mass_g ~ sex, color = ~ species, group = ~ species, stat = "summary")
```

The other two values computed by `mean_se()` are available (starting with 
`ggplot2` version 2.3) as `stat(ymin)` and `stat(ymax)`.

```{r, stats03, eval = utils::packageVersion("ggplot2") > "2.2.1"}
KidsFeet %>% 
  gf_jitter(length ~ sex, color = ~ domhand, width = 0.1, height = 0, alpha = 0.5) %>%
  gf_pointrange(length + stat(ymin) + stat(ymax) ~ sex, 
                color = ~ domhand, group = ~ domhand, stat="summary")
Penguins %>% 
  gf_jitter(body_mass_g ~ sex, color = ~ species, width = 0.1, height = 0, alpha = 0.3) %>%
  gf_pointrange(body_mass_g + stat(ymin) + stat(ymax) ~ sex, color = ~ species, 
                group = ~ species, size = 1.5, stat="summary")
```

Custom functions can be used by defining `fun.y`, `fun.ymin`, and `fun.ymax`,
or a single function `fun.data` that returns a data frame with variables 
named `y`, `ymin`, and `ymax`.


```{r, stats04, eval = utils::packageVersion("ggplot2") > "2.2.1"}
KidsFeet %>% 
  gf_point(length ~ sex, color = ~ domhand, width = 0.1, height = 0, alpha = 0.5,
           position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.3)) %>%
  gf_pointrange(length + stat(ymin) + stat(ymax) ~ sex, 
                color = ~ domhand, group = ~ domhand, stat="summary",
                fun.y = median, fun.ymin = min, fun.ymax = max,
                position = position_dodge(width = 0.6), size = 1.2)
```


### Plotting functions

`ggformula` provides two way to plot functions: `gf_function()` and `gf_fun()`.  They differ primarily in how one specifies the function to be plotted.  

 * `gf_function()` requires a (vectorized) function of one variable.
 * `gf_fun()` requires a formula describing how the y-values are computed from the x-values.
 
If these are used as a first layer, the limits for the x-axis must be specified.
 
```{r, functions}
gf_function(fun = sqrt, xlim = c(0, 10)) %>%
  gf_fun(2 + 3 * cos(10 * x) ~ x, color = "red", n = 800)
f <- makeFun(lm(totalbill ~ poly(month, 2), data = mosaicData::Utilities))
mosaicData::Utilities %>%  
  gf_point(totalbill ~ month, alpha = 0.6) %>%
  gf_fun(f(m) ~ m, color = "red")
```


## Chaining to create complex plots

### Multiple layers 

Often it is useful to overlay multiple layers onto a single plot.  This can be done 
by chaining them with `%>%`, the "then" operator from `magrittr`.  The `data` 
argument can be omitted if the new layers uses the same data as the first layer 
in the chain. 

The following plot illustrates how histograms and frequency polygons are related.

```{r, layers}
Penguins %>% 
  gf_histogram( ~ bill_length_mm, alpha = 0.3, fill = "navy", binwidth = 1) %>%
  gf_freqpoly( ~ bill_length_mm , binwidth = 1) %>% 
  gf_facet_grid(species ~ .)
```

A 2-d density plot can be augmented with a scatterplot.

```{r, layers02}
Penguins %>% 
  gf_density_2d(bill_length_mm ~ bill_depth_mm) %>%
  gf_jitter(bill_length_mm ~ bill_depth_mm, alpha = 0.2, width = 0.2, height = 0.2) %>%
  gf_facet_wrap(~species)
```

### Maps

Prior to `ggplot2` version 2.3,
basic maps could be created using `gf_polygon()`.

```{r, fig.width = 6, fig.height = 3, out.width = "75%"}
if (require(maps) && require(dplyr)) {
  US <- map_data("state") %>%
    dplyr::mutate(name_length = nchar(region))
  States <- US %>%
    dplyr::group_by(region) %>%
    dplyr::summarise(lat = mean(range(lat)), long = mean(range(long))) %>%
    dplyr::mutate(name = abbreviate(region, 3))

  gf_polygon(lat ~ long, data = US, group = ~ group,
             fill = ~ name_length, color = "white") %>%
  gf_text(lat ~ long, label = ~ name, data = States,
    color = "gray70", inherit = FALSE) %>%
  gf_refine(mosaic::theme_map())
}
```

Starting with `ggplot2` version 2.3, there is direct support for shape files.
This allows each row of a data frame to contain all the geometry information
for a region in a single column.  There is also better support for overlaying
information from multiple shape files.


```{r, fig.width = 6, fig.height = 3, out.width = "75%"}
if (require(maps) && require(dplyr) && require(sf) && require(purrr)) {
  USshape <- 
    sf::st_as_sf(maps::map('state', plot = FALSE, fill = TRUE)) %>%
    dplyr::mutate(
      name_length = nchar(as.character(ID)),
      centroid = purrr::map(geom, st_centroid),
      coords_x = purrr::map_dbl(centroid, 1),
      coords_y = purrr::map_dbl(centroid, 2)
    ) 
    
  gf_sf(fill = ~ factor(name_length), color = "white", data = USshape, alpha = 0.5) %>%
    gf_sf(data = sf::st_centroid(USshape), color = "white", alpha = 0.5, size = 3) %>%
    gf_text(coords_y ~ coords_x, label = ~ ID, color = "gray20", size = 2) %>%
    gf_labs(x = "", y = "") %>%
    gf_refine(coord_sf(), mosaic::theme_map()) %>%
    gf_labs(fill = "name length")
}
```


### New types of plots

`ggformula` adds some additional plot options to `ggplot2`

#### ASH plots

Average shifted histograms can be created with `gf_ash()`.  These plots 
average the height over all histograms with the same bin width.
ASH plots often work well with larger bin widths than you might use for an 
individual histogram.

```{r}
HELPrct %>% 
  gf_ash( ~ age, data = HELPrct, binwidth = 2) %>%
  gf_dhistogram( ~ age, binwidth = 2, alpha = 0.3) 
HELPrct %>% 
  gf_ash( ~ age, binwidth = 10) %>% 
  gf_dhistogram( ~ age, binwidth = 2, alpha = 0.3) 
```

#### Distribution plots

`gf_dist()` can be used to create plots of discrete and continuous distributions.


```{r}
gf_dist("pois", lambda = 5)
gf_dist("pois", lambda = 5, kind = "cdf")
gf_dist("gamma", shape = 3, rate = 4, geom = "area")
gf_dist("gamma", shape = 3, rate = 4, geom = "area", fill = ~ (x <= 1))
gf_dist("gamma", shape = 3, rate = 4, kind = "cdf")
```

When distribution parameters and plot attributes have the same name, 
there are two ways to avoid the name collision:

```{r}
# size is used by the binomial distribution functions and when plotting
gf_dist("binom", size = 20, prob = 0.25, plot_size = 5)
gf_dist("binom", params = list(size = 20, prob = 0.25), size = 3)
```



`ggformula` also provides an interface to `MASS::fitdistr()` for fitting distributions
to data and displaying the resulting pdf.

```{r}
x <- rgamma(1000, shape = 2, rate = 5)
gf_dhistogram( ~ x, alpha = 0.3) %>%
  gf_fitdistr(dist = "dnorm", color = ~ "Normal") %>% 
  gf_fitdistr(dist = "dgamma", color = ~ "Gamma") %>%
  gf_fitdistr(dist = "dweibull", color = ~ "Weibull") %>% 
  gf_labs(color = "")
```

### Global plot adjustments

There are a number of things we may want to do to the entire plot -- adjusting
labels, colors, fonts, etc.  `ggformula` provides wrappers to the `ggplot2` functions 
for this so that the chaining syntax can be used.

```{r, themes}
penguins %>% 
  gf_histogram( ~ body_mass_g, alpha = 0.2, fill = "navy",
              color = "gray50", binwidth = 200) %>%
  gf_freqpoly( ~ body_mass_g, binwidth = 200) %>%
  gf_facet_grid(species ~ .) %>% 
  gf_labs(x = "bill length (mm)", title = "Palmer Penguins") %>%
  gf_lims(x = c(2500, 6500)) %>% 
  gf_theme(theme = theme_minimal())

penguins %>% 
  gf_histogram( ~ body_mass_g | species ~ ., alpha = 0.5, fill = "white",
              binwidth = 200, color = "gray40") %>%
  gf_freqpoly( ~ body_mass_g, color = "skyblue", 
               binwidth = 200, size = 1.2, alpha = 0.8) %>%
  gf_labs(x = "bill length (mm)", title = "Palmer Penguins") %>%
  gf_lims(x = c(2500, 6500)) %>% 
  gf_theme(theme = theme_dark())
```

For convenience, a few modifications can also be made directly in the original
function call:

```{r}
penguins %>% 
  gf_dhistogram( ~ body_mass_g | species ~ ., 
                alpha = 0.5, fill = "skyblue", color = "navy", binwidth = 200, 
                xlab = "body mass (g)", title = "Palmer Penguins"
  ) 
```

## Horizontal Geoms

Some of the geoms in ggplot2 have a vertical orientation.
There are two ways to obtain horizontal versions of thse geoms:

* The ggstance package provides a set of horizontal geoms and positions;

* The ggplot2 package now provides an `orientation` argument that can be use to
obtain "native" horizontal or vertical geoms and positions.  

    Often ggplot2 will correctly determine which orientation is intended.  If not, set
`orientation = 'x'` or `orientation = 'y'` to explicitly set the direction of the geom.

ggformula supports both of these methods.

```{r}
gf_boxplot(high_temp ~ city, data = Weather)

# auto-detected because city is discrete
gf_boxplot(city ~ high_temp, data = Weather)

# auto-detected based on groups
gf_boxplot(month ~ high_temp | ~ city, group = ~ month, data = Weather)
gf_boxplot(high_temp ~ month | city ~ ., group = ~ month, data = Weather)

# defaults to veritcal without groups
gf_boxplot(month ~ high_temp | ~ city, data = Weather)

# explicit horizontal orientation
gf_boxplot(month ~ high_temp | city ~ ., data = Weather, orientation = 'y')

# using ggstance; notice the "h" at the end of gf_boxploth()
gf_boxploth(month ~ high_temp | city ~ ., data = Weather)
```